---
title: "MLE Simulation and Censored Data Analysis"
output: html_document
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)

```

# Simulate ~ N(0,1) data

```{r}
x <- rnorm(10^6)

ggplot(data.frame(x = x), aes(x = x)) + geom_histogram()
```
 
# Create a negative log-likelihood function for normal data

The negative Log-Likelihood function for a normal distribution is the negative of the sum of the log of the density function of the normal distribution.

```{r}
 rMLE <- function(params, y){
   
   mu = params[1]
   sigma = params[2]
   
   return(-1 * sum(dnorm(y, mean = mu, sd = sigma, log = TRUE)))
   
 }
``` 
 
# 100 percent observed MLE 

We use starting values kind of far away from true values to make sure the optimization converges to the right values for this easy case.

```{r}
 mle0 <- optim(par = c(2,3), rMLE, y = x)
 results <- mle0$par
 print(results)
```
 
Try with a different mean and standard deviation. Also, we start with starting values quite different to make sure the optimization converges to the right values for this easy case

```{r}
 x3 <- rnorm(10^6, mean = 3, sd = 6)
 
 mle3 <- optim(par = c(0,1), rMLE, y = x3)
 
 results3 <- mle3$par
 print(results3)
```

## Censored MLE function
 

```{r} 
 cMLE <- function(params, y.obs, y.cens = NULL){
   
   mu = params[1]
   sigma = params[2]
   
   
    obs_ll <- sum(dnorm(y.obs, mean = mu, sd = sigma, log = TRUE))
    cens_ll <- sum(pnorm(y.cens, mean = mu, sd = sigma, log.p = TRUE, lower.tail = TRUE))
    nll <- -1 * (obs_ll + cens_ll)
    
    return(nll)
   
 }
```
 
## Check that the censored ML is working. 

We will do  a loop by censoring more and more data below the MDL. The MLE estimates should deviate more and more from the observed mean and standard deviation as we censor more data.
 
```{r} 
MDL <- -1.5

p <- seq(0.1, 1, by = 0.1)

results <- list()
  
  
for (i in 1:length(p)){
  
  # random sample of p percent of the data
  idx <- c(1:length(x)) %in% sample.int(length(x), length(x) * p[i])
  
  cens_data <- rep(MDL, length(x) * (1-p[i]))
  obs_data <- x[idx]
  
  mle <- optim(par = c(0,1), cMLE, y.obs = obs_data, y.cens = cens_data)
  
  results[[i]] <- mle$par
  raw_mean=mean(c(obs_data,cens_data))
  
  print(paste("raw_mean=",raw_mean,"mu = ",results[[i]][1], "sd = ", results[[i]][2], "censored percentage = ", (100-p[i]*100)))
  
 # print(paste("mu = ",results[[i]][1], "sd = ", results[[i]][2], "censored percentage = ", (100-p[i]*100)))
  
}

```

The results seem to make sense; As the censoring decrease the MLE estimates get closer to the true values. Additionally, as the censoring increases, the mean moves farther towards -Inf, away from the MDL.

 
# MLE Trick for Strictly Positive Parameters 

We often have to estimate parameters that are strictly positive. For example, the mean concentration on the normal scale should be non-negative or a variance parameter (e.g. Kriging variance/sill).

One trick is to estimate the log of the parameter. Our estimate is then $exp(x)$. This way, the parameter is always positive. We could do constrained optimization, but the algorithms for unconstrained optimization are more efficient and more robust.


## MLE for observed data and log parameters
```{r}
 rMLE_log <- function(params, y){
   
   mu = exp(params[1])
   sigma = exp(params[2])
   
   return(-1 * sum(dnorm(y, mean = mu, sd = sigma, log = TRUE)))
   
 }

# Simulate strictly positive log-normal data
z <- rlnorm(10^3, meanlog = 3, sdlog = 1)

mle_log <- optim(par = c(5,1), rMLE_log, y = z)

results <- exp(mle_log$par)

print(paste("mu = ",results[1], "sd = ", results[2]))
print(paste("observed mean = ", mean(z), "observed sd = ", sd(z)))
```
# Combine functions fore censored MLE on Positive values

```{r}
cMLE_log <- function(params, y.obs, y.cens = NULL){
  mu = exp(params[1])
  sigma = exp(params[2])

  obs_ll <- sum(dnorm(y.obs, mean = mu, sd = sigma, log = TRUE))
    cens_ll <- sum(pnorm(y.cens, mean = mu, sd = sigma, log.p = TRUE, lower.tail = TRUE))
    nll <- -1 * (obs_ll + cens_ll)
    return(nll)
}

# Simulate strictly positive log-normal data
z <- rlnorm(10^3, meanlog = 3, sdlog = 1)


MDL <- 1.5
p <- seq(0.1, 1, by = 0.1)

results <- list()

for (i in 1:length(p)){
  
  # random sample of p percent of the data
  idx <- c(1:length(z)) %in% sample.int(length(z), length(z) * p[i])
  
  cens_data <- rep(MDL, length(z) * (1-p[i]))
  obs_data <- z[idx]
  
  mle <- optim(par = c(mean(z),sd(z)), cMLE_log, y.obs = obs_data, y.cens = cens_data)
  
  results[[i]] <- exp(mle$par)
  
  raw_mean=mean(c(obs_data,cens_data))
  
  print(paste("raw_mean=",raw_mean,"mu = ",results[[i]][1], "sd = ", results[[i]][2], "censored percentage = ", (100-p[i]*100)))
  
}

print(paste("observed mean = ", mean(z), "observed sd = ", sd(z)))  
 

```

